---
title: "2020 SEM1 STATS 765 project code"
author: "Ian Chen"
date: "10/06/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, cache=TRUE}
# Libraries
library(readxl)
```

# Loading the datasets

```{r loading the datasets, cache=TRUE}
# General electorates
ge2014 <- read.csv("general-electoral-district-2014.csv")
ge2014 <- ge2014[order(ge2014$GED2014_V1_00), -1]
rownames(ge2014) <- ge2014$GED2014_V1_00
maori2014 <- read.csv("maori-electoral-district-2014.csv")
maori2014 <- maori2014[, -1]
ge2020 <- read.csv("general-electorates-2020-data.csv")
colnames(ge2020)[1] <- "GED2020_V1_00"
maori2020 <- read.csv("maori-electorates-2020-data.csv")
colnames(maori2020)[1] <- "MED2020_V1_00"
head(ge2014)[1:4]

# Meshblocks
meshblocks2013 <- read.csv("meshblock-boundaries-2013.csv")
meshblocks2013 <- meshblocks2013[, -1]
meshblocks2020 <- read.csv("meshblock-2020-generalised.csv")
meshblocks2020 <- meshblocks2020[, -1]
head(meshblocks2020)[1:4]

# CLASSIFICATION OF MESHBLOCKS

# Meshblocks for various years with miscellaneous information
mbclass <- read.csv("geographic-areas-file-2020.csv")
colnames(mbclass)[1] = "MB2020_code"

# Meshblock labels over census and election years
yearlymbs <- read_excel("MB-only_census-and-electorate-years.xlsx")
# Meshblock labels 2014 election
mbelecs2014 <- read_excel("MB-and-electorates.xlsx")
colnames(mbelecs2014)[c(6, 7)] = c("2014_AREA_SQ_KM", "2014_LAND_AREA_SQ_KM")
# Meshblock labels 2008 election
mbelecs2008 <- read_excel("MB-and-electorates.xlsx", sheet = "MB08_GED07_MED07")
colnames(mbelecs2008)[c(6, 7)] = c("2008_AREA_SQ_KM", "2008_LAND_AREA_SQ_KM")

# Meshblocks 2020 with corresponding electorate allocation
mbelecpop2020 <- read.csv("meshblock-electoral-populations-2020-data (2).csv")
colnames(mbelecpop2020)[c(1, 10, 11)] = c("MB2020_V2_00", "2020_LAND_AREA_SQ_KM", "2020_AREA_SQ_KM")
# Converting the meshblock numbers to strings with leading zeroes
mbelecpop2020[, 1] = sprintf("%07d", mbelecpop2020[, 1])

# Party percentage votes 2014
partperc2014 <- read.csv("e9_part5_(2014).csv", skip = 2)
colnames(partperc2014) = c("Electoral District", "ACT NEW ZEALAND Votes",
                           "ACT NEW ZEALAND Percentage", "GREEN PARTY Votes",
                           "GREEN PARTY Percentage", "LABOUR PARTY Votes", "LABOUR PARTY Percentage",
                           "MĀORI PARTY Votes", "MĀORI PARTY Percentage", "NATIONAL PARTY Votes",
                           "NATIONAL PARTY Percentage", "NEW ZEALAND FIRST PARTY Votes",
                           "NEW ZEALAND FIRST PARTY Percentage", "UNITED FUTURE PARTY Votes",
                           "UNITED FUTURE PARTY Percentage", "OTHER Votes", "OTHER Percentage",
                           "TOTAL Votes")
# Party percentage votes 2017
partperc2017 <- read.csv("percentage-votes-for-registered-parties_(2017).csv", skip = 2)
colnames(partperc2017) = c("Electoral District", "ACT NEW ZEALAND Votes",
                           "ACT NEW ZEALAND Percentage", "GREEN PARTY Votes",
                           "GREEN PARTY Percentage", "LABOUR PARTY Votes", "LABOUR PARTY Percentage",
                           "NATIONAL PARTY Votes", "NATIONAL PARTY Percentage",
                           "NEW ZEALAND FIRST PARTY Votes", "NEW ZEALAND FIRST PARTY Percentage",
                           "OTHER Votes", "OTHER Percentage", "TOTAL Votes")
partperc2017$`Electoral District` <- as.character(partperc2017$`Electoral District`)
# Converting electorate names to ASCII
partperc2017[20, 1] = "Kaikoura"
partperc2017[23, 1] = "Mangere"
partperc2017[36, 1] = "Ohariu"
partperc2017[37, 1] = "Otaki"
partperc2017[43, 1] = "Rangitikei"
partperc2017[49, 1] = "Tamaki"
partperc2017[51, 1] = "Taupo"
partperc2017[53, 1] = "Te Atatu"
partperc2017[67, 1] = "Ikaroa-Rawhiti"
partperc2017[68, 1] = "Tamaki Makaurau"
partperc2017[69, 1] = "Te Tai Hauauru"
partperc2017[73, 1] = "Maori Electorate Totals"

head(partperc2017)[1:3]
```

# Data Exploration

```{r dataframe combining the meshblock data, cache=TRUE}
# Creating a custom dataframe to compare meshblock changes over election years
mbcomp <- merge(yearlymbs[, c(1, 3, 5)], mbelecpop2020[, c(1, 6, 9, 10, 11)])
mbcomp <- merge(mbcomp, mbelecs2014[, c(1, 3, 5, 6, 7)])
mbcomp <- merge(mbcomp, mbelecs2008[, c(1, 3, 5, 6, 7)])
mbcomp[1,]
```

# Electorate changes

Voter percentage is projected to be 81.25% of the total eligible population.

Because population values at the electorate and meshblock levels were not released publicly until 2020, the data is missing for 2014. Therefore, it should not be used as a metric for measuring the change in electorate boundaries' effect on the voter population.

A metric that exists for both 2014 and 2020 is electorate and meshblock land area, by $km^2$. Land area is used, because it is not fair to assume that the population of New Zealand has a presence in meshblocks of high water area. It is possible that this is the case (those living on offshore platforms at the time of the census), however these individuals are outliers and not representative of the vast majority of the New Zealand population. Thus, land area is used in favour of total area.

We assume, for this exercise, that the population is uniformly distributed across the land area of each electorate of New Zealand.


```{r general algorithm, cache=TRUE}
elecs2014 <- partperc2017[, 1]
elecs2020 <- c("Auckland Central", "Banks Peninsula", "Bay of Plenty", "Botany", "Christchurch Central", "Christchurch East", "Coromandel", "Dunedin", "East Coast", "East Coast Bays", "Epsom", "Hamilton East", "Hamilton West", "Hutt South", "Ilam", "Invercargill", "Kaikoura", "Kaipara ki Mahurangi", "Kelston", "Mana", "Mangere", "Manurewa", "Maungakiekie", "Mt Albert", "Mt Roskill", "Napier", "Nelson", "New Lynn", "New Plymouth", "North Shore", "Northcote", "Northland", "Ohariu", "Otaki", "Pakuranga", "Palmerston North", "Panmure-Otahuhu", "Papakura", "Port Waikato", "Rangitata", "Rangitikei", "Remutaka", "Rongotai", "Rotorua", "Selwyn", "Southland", "Taieri", "Takanini", "Tamaki", "Taranaki-King Country", "Taupo", "Tauranga", "Te Atatu", "Tukituki", "Upper Harbour", "Waikato", "Waimakariri", "Wairarapa", "Waitaki", "Wellington Central", "West Coast-Tasman", "Whanganui", "Whangaparaoa", "Whangarei", "Wigram", "General Electorate Totals", "Hauraki-Waikato", "Ikaroa-Rawhiti", "Tamaki Makaurau", "Te Tai Hauauru", "Te Tai Tokerau", "Te Tai Tonga", "Waiariki", "Maori Electorate Totals", "Combined Totals")

predvotes <- data.frame(matrix(NA, nrow = 75, ncol = 14))
colnames(predvotes) = colnames(partperc2017)
predvotes[, 1] <- elecs2020

# Algorithm for General Electorates
for (i in 1:65) {
  # Incoming meshblocks
  incoming <- mbcomp[which(mbcomp$GED2020_V1_00_NAME_ASCII == ge2020[i, 3]), ]
  uniques <- unique(incoming[, 8])
  elecindices <- c(0)
  for (o in 1:length(uniques)) {
    elecindices[o] <- which(elecs2014 == uniques[o])
  }
  parts <- partperc2017[elecindices, ]
  voter_pop <- c(NA)
  for (j in 1:nrow(parts)) {
    sa <- sum(incoming[which(incoming[, 8] == parts[j, 1]), ]$`2020_LAND_AREA_SQ_KM`)
    voters <- round(ge2020[i, 4] * 0.8125 * sa / ge2020[i, 5])
    if (j == 1) {
      voter_pop <- c(voters)
    } else {
      voter_pop <- c(voter_pop, voters)
    }
  }
  # Calculating new voter numbers with same percentage votes
  for (k in 1:nrow(parts)) {
    total <- 0
    for (l in c(3, 5, 7, 9, 11, 13)) {
      parts[k, l - 1] = round(voter_pop[k] * (parts[k, l] / 100))
      total <- total + parts[k, l - 1]
    }
    parts[k, 14] = total
  }
  # Setting up the row for the 2020 electorate
  parts[nrow(parts) + 1, ] = NA
  # Calculating totals for the 2020 electorate
  for (m in c(2, 4, 6, 8, 10, 12, 14)) {
    parts[nrow(parts), m] <- sum(parts[, m], na.rm = TRUE)
  }
  # Calculating the new percentage votes for each party
  for (n in c(3, 5, 7, 9, 11, 13)) {
    parts[nrow(parts), n] <- round(parts[nrow(parts), n - 1] / parts[nrow(parts), 14] * 100, 2)
  }
  predvotes[i, 2:14] <- parts[nrow(parts), 2:14]
}

# Algorithm for Maori Electorates
for (i in 1:7) {
  # Incoming meshblocks
  incoming <- mbcomp[which(mbcomp$MED2020_V1_00_NAME_ASCII == maori2020[i, 3]), ]
  uniques <- unique(incoming[, 9])
  elecindices <- c(0)
  for (o in 1:length(uniques)) {
    elecindices[o] <- which(elecs2014 == uniques[o])
  }
  parts <- partperc2017[elecindices, ]
  voter_pop <- c(NA)
  for (j in 1:nrow(parts)) {
    sa <- sum(incoming[which(incoming[, 9] == parts[j, 1]), ]$`2020_LAND_AREA_SQ_KM`)
    voters <- round(maori2020[i, 4] * 0.8125 * sa / maori2020[i, 5])
    if (j == 1) {
      voter_pop <- c(voters)
    } else {
      voter_pop <- c(voter_pop, voters)
    }
  }
  # Calculating new voter numbers with same percentage votes
  for (k in 1:nrow(parts)) {
    total <- 0
    for (l in c(3, 5, 7, 9, 11, 13)) {
      parts[k, l - 1] = round(voter_pop[k] * (parts[k, l] / 100))
      total <- total + parts[k, l - 1]
    }
    parts[k, 14] = total
  }
  # Setting up the row for the 2020 electorate
  parts[nrow(parts) + 1, ] = NA
  # Calculating totals for the 2020 electorate
  for (m in c(2, 4, 6, 8, 10, 12, 14)) {
    parts[nrow(parts), m] <- sum(parts[, m], na.rm = TRUE)
  }
  # Calculating the new percentage votes for each party
  for (n in c(3, 5, 7, 9, 11, 13)) {
    parts[nrow(parts), n] <- round(parts[nrow(parts), n - 1] / parts[nrow(parts), 14] * 100, 2)
  }
  predvotes[i + 66, 2:14] <- parts[nrow(parts), 2:14]
}
```

The above algorithms calculate approximately 81.25% of the electorate populations for 57 electorates. All Maori electorates were calculated accurately. However, 15 general electorates have voter populations that are either non-existent or not within 0.1% of the 81.25% voter population. Because I'm lazy, I will manually calculate them (as opposed to trying to find the issue in the algorithm that causes these 15 general electorates to not work properly).

EDIT: Turns out a slightly altered algorithm works, despite the fact that it is almost identical to the first one. The change was made to the original algorithms.

```{r the problematic electorates, cache=TRUE}
# The function that fixes the original algorithm
for (i in c(6, 8, 15, 18, 23, 28, 32, 37, 38, 39, 40, 46, 48, 56, 63)) {
  incoming <- mbcomp[which(mbcomp$GED2020_V1_00_NAME_ASCII == ge2020[i, 3]), ]
  # The Altered section
  uniques <- unique(incoming[, 8])
  elecindices <- c(0)
  for (o in 1:length(uniques)) {
    elecindices[o] <- which(elecs2014 == uniques[o])
  }
  parts <- partperc2017[elecindices, ]
  # The altered section ends here
  voter_pop <- c(NA)
  for (j in 1:length(uniques)) {
    sa <- sum(incoming[which(incoming[, 8] == parts[j, 1]), ]$`2020_LAND_AREA_SQ_KM`)
    voters <- round(ge2020[i, 4] * 0.8125 * sa / ge2020[i, 5])
    if (j == 1) {
      voter_pop <- c(voters)
    } else {
      voter_pop <- c(voter_pop, voters)
    }
  }
  for (k in 1:nrow(parts)) {
    total <- 0
    for (l in c(3, 5, 7, 9, 11, 13)) {
      parts[k, l - 1] = round(voter_pop[k] * (parts[k, l] / 100))
      total <- total + parts[k, l - 1]
    }
    parts[k, 14] = total
  }
  parts[nrow(parts) + 1, ] = NA
  for (m in c(2, 4, 6, 8, 10, 12, 14)) {
    parts[nrow(parts), m] <- sum(parts[, m], na.rm = TRUE) 
  }
  for (n in c(3, 5, 7, 9, 11, 13)) {
    parts[nrow(parts), n] <- round(parts[nrow(parts), n - 1] / parts[nrow(parts), 14] * 100, 2)
  }
  predvotes[i, 2:14] <- parts[nrow(parts), 2:14]
}
```


Completing the dataframe with totals and percentage votes of total:

```{r totalling everything, cache=TRUE}
# Summing the total votes
for (p in c(14, 2, 4, 6, 8, 10, 12)) {
  predvotes[66, p] <- sum(predvotes[, p][1:65])
  predvotes[74, p] <- sum(predvotes[, p][67:73])
  predvotes[75, p] <- predvotes[66, p] + predvotes[74, p]
  # Calculating the percentage of the total electorate vote and total country vote
  if (p != 14) {
    predvotes[66, p + 1] <- round(predvotes[66, p] / predvotes[66, 14] * 100, 2)
    predvotes[74, p + 1] <- round(predvotes[74, p] / predvotes[74, 14] * 100, 2)
    predvotes[75, p + 1] <- round(predvotes[75, p] / predvotes[75, 14] * 100, 2)
  }
}
```


# Looking at the results

My predicted party votes per electorate showed that:
      1) Overall, electorate boundary changes do not have an impactful effect on the party vote. Most boundary changes were very minor, and given that this model only takes the land surface area changes (assuming a uniform distribution of New Zealand's population across the land area), this is to be expected.
      2) The overall percentage party vote changes from 2017 are: ACT -0.01%, Green -0.17%, Labour +0.62%, National -1.04%, NZ First +0.10%, Other +0.51%. The Labour Party stands to gain the most from the electoral boundary changes, while the National Party stands to lose the most from the changes. This is, however, not a large enough gain or loss to alter the popular party vote.
      3) As a result of the boundaries having very little effect, the popular vote still goes to the National Party, which would imply that Labour would need to form a coalition in order to retain their ministry. However, 2020 has been a very strange year thus far, and anecdotal evidence points to the popular opinion that current Prime Minister Jacinda Ardern has coped well with the various crises New Zealand has suffered over the past year (Christchurch mosque shootings, White Island eruption, COVID-19). Latest polls post-COVID-19 lockdown point to the National Party having historically low opinion ratings, as well as the sacking of then-National leader Simon Bridges from the leadership position in favour of Todd Muller (who himself is proving to be somewhat unpopular) not too long ago, which does not bode well for their campaign. The eye test points to Labour winning the popular vote by a landslide, and (my 2c) I would expect Labour to gain a large enough advantage over National that they can dissolve the current coalition and rule as a single party government (if that's even how it works, and if that's even what Labour wants, as opposed to keeping the coalition to maintain the current cabinet for continuity's sake).
      4) Voting population multiplies by a factor of $\frac{3819219}{2591896} = 1.4735$, an increase of almost 50% since the previous election. This is a strange value, as the voting population was expected to linearly increase by about 5% overall, and yet the model projects the voting population to increase by 10x that. The 2017 voter numbers seem slightly off, as the total voting population in 2017 does not make sense as ~77% of New Zealand's population (as New Zealand's population is widely considered to be in the range of 4.5-5 million people - only very recently is New Zealand's population estimated to be over 5 million).